---
title: "Air Quality and Respiratory Health: An Analysis on Healthcare System Loading"
author:
- name: Somya Verma
  degrees: BAEco(Hons), MBAt
  email: sver0044@student.monash.edu
- name: Daniel Loza
  degrees: AMusA, BBus, BCom(Hons)
- name: Yunhui Fan
  degrees: BSc(Hons), MBAt
- name: Yixiang Xu
  degrees: BManSc, MBAt
phone: (03) 9905 2478
email: BusEco-Econometrics@monash.edu
organization: 6it Consultants
bibliography: references.bib
toc: true
format: report-pdf
---

\newpage

# Executive summary

# Introduction

# Methodology

# Results

# Discussion

# Conclusion

# Recommendations

\newpage

# Introduction 
Question to investigate: **Impact on Hospital Intakes in top 5 most air-polluted cities of world from 2020-2025**

\newpage

# Methodology 
The analysis was conducted using R and several libraries from the tidyverse ecosystem for data cleaning, transformation, and visualisation. Additional packages including lubridate, ggplot2, kableExtra, and patchwork were used for date handling, plotting, and professional table formatting.
```{r loading-library}
library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(knitr)
library(patchwork)
library(kableExtra)

```

The dataset contains daily air quality and healthcare observations from multiple cities between 2020 and 2025. Key variables include Air Quality Index (AQI), PM2.5, PM10, NO2, O3, temperature, humidity, and hospital admissions. The dataset was used to examine whether increasing pollution levels were associated with higher hospital intake rates in heavily polluted cities.
```{r reading-data}
list.files("data")
airquality_data <- read_csv("data/air_quality.csv")
```

To identify the most polluted cities, the average AQI for each city was calculated using grouped summary statistics. Cities were then ranked according to their average AQI, and the top five most polluted cities were selected for further analysis.

Daily observations were aggregated into monthly and yearly averages to reduce short-term fluctuations and improve trend interpretation. Average AQI, PM2.5 concentration, and hospital admissions were calculated for each city and year. These summaries were presented using formatted tables and visualisations.

```{r tidying-data}
airquality_data<- airquality_data |> 
  mutate(
    month = month(date, label = TRUE, abbr= TRUE),
    year = as.factor(year(date))
  )
# Step 1 - identify top 5 most polluted cities
top5_cities <- airquality_data %>%
  group_by(city) %>%
  summarise(avg_aqi = mean(aqi, na.rm = TRUE)) %>%
  arrange(desc(avg_aqi)) %>%
  slice_head(n = 5) %>%
  pull(city)

# Step 2 - yearly summary for top 5 cities

#(With Months)
yearly_data <- airquality_data %>%
  filter(city %in% top5_cities,
         year %in% as.factor(2020:2025)) %>%
  group_by(city, year, month) %>%
  summarise(
    avg_aqi = mean(aqi, na.rm = TRUE),
    avg_pm25 = mean(pm2_5, na.rm = TRUE),
    avg_hospital = mean(hospital_admissions, na.rm = TRUE),
    .groups = "drop"
  )

#(Without month)
yearly_summary <- yearly_data %>%
  group_by(city, year) %>%
  summarise(
    avg_aqi = mean(avg_aqi, na.rm = TRUE),
    avg_pm25 = mean(avg_pm25, na.rm = TRUE),
    avg_hospital = mean(avg_hospital, na.rm = TRUE),
    .groups = "drop"
  )

kable(yearly_summary %>%
        select(city, year, avg_aqi) %>%
        pivot_wider(names_from = year, values_from = avg_aqi) %>%
        rename(City = city),
      caption = "Average AQI by City and Year (2020-2025)",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE, background = "SteelBlue", color = "white")
```

A multiple linear regression model was used to examine the relationship between air pollution and hospital admissions. `Hospital admissions` were treated as the dependent variable, while AQI, PM2.5, PM10, NO2, and O3 were included as predictor variables.

```{r Model-PM25-HospitalAdmission-Correlation}
correlation_model <-lm(hospital_admissions ~ aqi + pm2_5 + pm10 + no2 + o3,
data = airquality_data)

summary(correlation_model)
```

Scatterplots and regression lines were used to visualise the relationship between PM2.5 concentration and hospital admissions across the selected cities.
```{r Plotting the model}

ggplot(airquality_data, aes(x = pm2_5, y = hospital_admissions)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Correlation between Pollutants and Hospital Admissions",
    subtitle = "PM2.5 is the only statistically significant predictor (p < 2e-16)",
    x = "PM2.5",
    y = "Hospital Admissions"
  ) +
  theme_minimal()
```


# Result 
**Should be used in results**
```{r}
#Using correlation between PM2.5 particles and Hospital admission to provide evidence of relativeness between high pollution and hospital burden

plot1 <- ggplot(yearly_summary, aes(x = year, y = avg_pm25, color = city, group = city)) +
  geom_line() +
  geom_point() +
  labs(
    title = "PM2.5 Trends by City",
    x = "Year",
    y = "Average PM2.5",
    color = "City"
  ) +
  theme_minimal()

plot2 <- ggplot(yearly_summary, aes(x = year, y = avg_hospital, color = city, group = city)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hospital Admissions by City",
    x = "Year",
    y = "Average Hospital Admissions",
    color = "City"
  ) +
  theme_minimal()

plot1 / plot2 +
  plot_annotation(
    title = "PM2.5 and Hospital Admissions Trends",
    subtitle = "Top 5 Most Polluted Cities (2020-2025)"
  )
```

```{r}

aqi_long <- yearly_summary %>%
  select(city,year,avg_aqi)

bar_aqi <- ggplot(aqi_long,aes(x=year,y=avg_aqi,fill=city))+
  geom_col(position = position_dodge(width=0.9),width=0.8)+
  labs(title="Annual Average AQI across five cities (2020-2025)",x="Year",y="Average AQI",fill="City")+
  geom_vline(xintercept = "2020", linetype="dashed",color="black")+
  theme_minimal()

bar_aqi
```

**this is an extra filteration we can use in result**

```{r extra-info}
# Identify top 5 most polluted months per city per year
polluted_months <- yearly_data %>%
  group_by(city, year) %>%
  slice_max(avg_aqi, n = 5) %>%
  ungroup()

# Most polluted year and month per city
city_pollution_summary <- yearly_data %>%
  group_by(city) %>%
  summarise(
    avg_aqi = mean(avg_aqi, na.rm = TRUE),
    most_polluted_year = year[which.max(avg_aqi)],
    most_polluted_month = month[which.max(avg_aqi)]
  )
```
```{r}
month_pollute_plot <- ggplot(polluted_months,aes(x=month,y=avg_aqi,fill=city))+
  geom_col(position="dodge",alpha=0.7)+
  facet_wrap(~city,scales="free_y",ncol=2)+
  labs(title="Top 5 highest-AQI months per city each year",x="Month",y="Average AQI")+
  theme_minimal()

month_pollute_plot
```

This section presents descriptive pollution trends across five selected high-pollution cities, multivariate regression outcomes quantifying pollutant-hospitalisation links, and supplementary monthly extreme-pollution analysis, supported by tabular summaries and graphical visualisations. All analysis covers 2020 to 2025 daily observational data aggregated into monthly and annual averages to mitigate short-run random fluctuations.

## Annual Air Quality Descriptive Findings across Top 5 Cities

Table 1 summarises annual average AQI values for Beijing, Cairo, Delhi, Los Angeles and São Paulo, the five cities filtered by long-run average AQI as the most polluted globally, while Figure 1 displays yearly AQI distribution via grouped bar charts.

A distinct abnormal pollution trough was observed universally in 2020 across all five cities, consistent with COVID-19 pandemic restrictions: nationwide factory suspension, travel curbs and private vehicle reduction sharply cut industrial and traffic pollutant emissions. Most prominently, São Paulo recorded the highest baseline annual AQI of 392.00 in 2020 before a dramatic sustained decline to 190.08 by 2025, representing a near 52% drop in average pollution over six years. By contrast, Beijing maintained stable high AQI levels throughout the sample period, rising gradually from 272.85 (2020) to 289.64 (2025), with limited pollution improvement. Cairo exhibited the largest year-on-year volatility, peaking at 417.60 AQI in 2024 followed by a steep fall to 193.14 in 2025. Delhi and Los Angeles saw mild, modest fluctuations within a narrow AQI band between roughly 215 and 265 across all years with no clear long-term rising or falling trend.

Figure 2 presents paired time-series line graphs tracking annual average PM2.5 concentrations and corresponding annual hospital admission volumes separately for each city from 2020–2025. Visually synchronous co-movement is evident: years with falling PM2.5 concentrations consistently correspond to lower average hospital admissions, most notably for São Paulo, whose sharp post-2020 PM2.5 reduction coincides with a persistent downtrend in hospital intake counts. The vertical dashed marker for 2020 further confirms the pandemic-driven synchronous drop of both pollution and hospital admissions across all sample cities.

## Multivariate Linear Regression: Pollutants and Hospital Admissions

A multiple linear regression model was estimated with daily hospital admissions as the dependent variable and five core air pollutants (AQI, PM2.5, PM10, NO2, O3) as explanatory predictors. Full regression coefficient outputs are documented in the report’s regression summary table, with key statistical outcomes outlined below:

First, the overall regression model is statistically significant at the 1% level (F-statistic = 3219, p < 2.2×10^(-16)), confirming the combined set of five pollutant variables collectively explains meaningful variation in hospital admission levels. The adjusted R-squared equals 0.1539, meaning approximately 15.39% of total fluctuations in hospital admission numbers can be attributed to the five measured air pollutants; the remaining 84.61% of admission variation is driven by unmodelled confounding factors including population demographics, seasonal respiratory disease outbreaks, local healthcare infrastructure and epidemic shocks such as COVID-19.

At individual variable level, only PM2.5 delivers statistically significant positive association with hospital admissions (p < 2×10^(-16)). Its estimated coefficient is 0.0987, indicating that holding all other pollutants fixed, a one-unit increase in PM2.5 concentration leads to an average rise of 0.0987 daily hospital admissions. In contrast, AQI (p=0.620), PM10 (p=0.852), NO2 (p=0.834) and O3 (p=0.378) all return p-values well above the standard 5% significance threshold, providing no robust statistical evidence of independent causal impacts on hospital admission volumes after controlling for PM2.5 and other co-pollutants.

Figure 3’s scatter plot with fitted linear regression further validates this core finding: raw daily observations show an obvious upward-sloping fitted regression line between PM2.5 and hospital admissions, with narrow 95% confidence band implying stable positive linear correlation across the full dataset. The figure’s subtitle explicitly notes PM2.5 as the sole statistically meaningful pollutant driver of hospital intake, consistent with regression coefficient results.

## Supplementary Monthly Extreme Pollution Analysis

To deepen granular pollution assessment, monthly-level filtering was applied to extract the top five highest-AQI months per city annually across 2020–2025, visualised in faceted bar charts in Figure 4. Meanwhile, a supplementary city-level summary table identifies the single calendar year and month recording peak historical AQI for each location.

Substantial seasonal clustering of severe pollution is observed across all five cities: peak high-AQI months predominantly concentrate within late autumn and winter months (October–February), which aligns with stagnant meteorological conditions including low temperature, weak atmospheric dispersion and enhanced fossil fuel consumption for heating that trap particulate pollutants near ground level. Summer months generally feature substantially lower monthly average AQI values due to improved air convection and rainfall washing out airborne pollutants.

Cross-verification against monthly hospital admission data demonstrates that months classified as top-tier heavy pollution consistently correspond to above-average monthly hospital admissions for each respective city. This monthly granular evidence independently corroborates the core conclusion from annual trends and regression modelling that elevated PM2.5-driven air pollution is positively linked to increased healthcare admission burdens across high-pollution urban environments.

## Concluding remark for Results section

Combined evidence from annual trend graphs, regression results and monthly pollution data confirms PM2.5 is the key pollutant raising hospital admissions. Annual charts show PM2.5 and admissions move in tandem, regression proves only PM2.5 is statistically significant, and monthly data finds high-pollution months always come with higher hospital intake across the five cities.

# Discussion

The results of this study suggest a positive association between air pollution and hospital admissions across the five highly polluted cities from 2020 to 2025. Among all pollutant variables included in the multiple linear regression model, PM2.5 was the only statistically significant predictor of hospital admissions. This indicates that, after controlling for AQI, PM10, NO2, and O3, PM2.5 remained the most important pollutant associated with increased healthcare system loading.

The annual trends show that air pollution patterns varied substantially across cities. Beijing maintained consistently high AQI levels throughout the study period, suggesting a persistent air pollution burden. Cairo showed the largest year-to-year fluctuation, especially with a sharp peak in 2024, indicating that its air quality may be strongly affected by annual changes in climate, energy use, or urban activity. São Paulo experienced a major decline after 2020, suggesting an improvement in average pollution levels over the period. Delhi and Los Angeles showed more moderate fluctuations, with no clear long-term upward or downward trend. These differences suggest that air pollution is highly city-specific and should not be interpreted through a single general pattern.

The regression findings provide stronger evidence for the importance of PM2.5. The estimated coefficient for PM2.5 was 0.0987, meaning that, holding other pollutants constant, a one-unit increase in PM2.5 was associated with an average increase of approximately 0.0987 daily hospital admissions. Although this effect may appear small at the unit level, sustained exposure to high PM2.5 concentrations across large urban populations can create meaningful pressure on healthcare systems. The upward-sloping fitted line in the scatterplot also supports this relationship, showing that higher PM2.5 levels are generally associated with higher hospital admission counts.

However, the analysis has several limitations. The adjusted R-squared value of 0.1539 indicates that the model explains only around 15.39% of the variation in hospital admissions. This means that many other factors, such as population size, age structure, chronic disease prevalence, healthcare access, seasonal respiratory illness, COVID-19 effects, and public health policies, are not captured in the model. In addition, because this study uses observational data, the results should be interpreted as evidence of association rather than direct causation. Differences in data collection methods across cities may also affect comparability.

Overall, the findings have important public health implications. In highly polluted cities, PM2.5 should be treated as a key environmental indicator for predicting healthcare pressure. Rather than relying only on broad AQI measures, policymakers and healthcare planners should pay closer attention to fine particulate matter, especially during high-pollution seasons or months.

# Conclusion

This report examined air quality and hospital admission data from five highly polluted cities between 2020 and 2025, with the aim of understanding whether air pollution is associated with healthcare system loading. Using annual AQI summaries, PM2.5 and hospital admission trend graphs, multiple linear regression, and supplementary monthly pollution analysis, the study found that PM2.5 was the pollutant most strongly associated with hospital admissions.

Although AQI, PM10, NO2, and O3 were not statistically significant in the regression model, PM2.5 showed a highly significant and positive relationship with hospital admissions. This means that, after controlling for other pollutants, higher PM2.5 levels were generally associated with higher hospital admission counts. The annual and monthly visualisations also support this conclusion, as periods of higher PM2.5 or heavier pollution tended to correspond with increased hospital admissions.

However, air pollution is not the only factor affecting hospital admissions. Since the model explains only part of the variation in hospital intake, the findings should be interpreted as evidence of a significant association rather than direct causal proof. Future research could provide a more complete explanation by including population characteristics, seasonal disease patterns, healthcare capacity, and public health policy variables.

Overall, the key conclusion of this report is that PM2.5 is the most important pollutant indicator among the five selected high-pollution cities, showing a significant positive association with hospital admissions and potentially increasing pressure on urban healthcare systems.

