---
title: "Air Quality and Respiratory Health: An Analysis on Healthcare System Loading"
author:
- name: Somya Verma
  degrees: BAEco(Hons), MBAt
  email: sver0044@student.monash.edu
- name: Daniel Loza
  degrees: AMusA, BBus, BCom(Hons)
- name: Yunhui Fan
  degrees: BSc(Hons), MBAt
- name: Yixiang Xu
  degrees: BManSc, MBAt
phone: (03) 9905 2478
email: BusEco-Econometrics@monash.edu
organization: 6it Consultants
bibliography: references.bib
toc: true
format: report-pdf
---

\newpage

# Executive summary

# Introduction

Question to investigate: **Impact on Hospital Intakes in top 5 most air-polluted cities of world from 2020-2025**

# Methodology

The analysis was conducted using R and several libraries from the tidyverse ecosystem for data cleaning, transformation, and visualisation. Additional packages including lubridate, ggplot2, kableExtra, and patchwork were used for date handling, plotting, and professional table formatting.
```{r loading-library}
library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(knitr)
library(patchwork)
library(kableExtra)
library(broom)

```

The dataset contains daily air quality and healthcare observations from multiple cities between 2020 and 2025. Key variables include Air Quality Index (AQI), PM2.5, PM10, NO2, O3, temperature, humidity, and hospital admissions. The 8 cities dataset was used to examine whether increasing pollution levels were associated with higher hospital intake rates in most and least polluted cities.
```{r reading-data}
list.files(here("data"))
airquality_data <- read_csv(here("data/air_quality.csv"))
```

To identify the most polluted cities, the average AQI for each city was calculated using grouped summary statistics. Cities were then ranked according to their average AQI, and the top five most polluted cities were selected for further analysis.

Daily observations were aggregated into monthly and yearly averages to reduce short-term fluctuations and improve trend interpretation. Average AQI, PM2.5 concentration, and hospital admissions were calculated for each city and year. These summaries were presented using formatted tables and visualisations.

```{r tidying-data}
airquality_data<- airquality_data |> 
  mutate(
    month = month(date, label = TRUE, abbr= TRUE),
    year = as.factor(year(date))
  )
```

```{r classification-of-data}
# Step 2 — Club all daily data into monthly summaries

monthly_summary <- airquality_data |>

  group_by(city, year, month) |>

  summarise(
    avg_aqi = mean(aqi, na.rm = TRUE),
    avg_pm25 = mean(pm2_5, na.rm = TRUE),
    avg_pm10 = mean(pm10, na.rm = TRUE),
    avg_no2 = mean(no2, na.rm = TRUE),
    avg_o3 = mean(o3, na.rm = TRUE),

    avg_temperature =
      mean(temperature, na.rm = TRUE),

    avg_humidity =
      mean(humidity, na.rm = TRUE),

    total_hospital =
      sum(hospital_admissions, na.rm = TRUE),

    avg_hospital_capacity =
      mean(hospital_capacity, na.rm = TRUE),

    .groups = "drop"
  )

# Step 3 — Club all months into yearly summaries

yearly_summary <- monthly_summary |>

  group_by(city, year) |>
    summarise(
    avg_aqi = mean(avg_aqi, na.rm = TRUE),
    avg_pm25 = mean(avg_pm25, na.rm = TRUE),
    avg_pm10 = mean(avg_pm10, na.rm = TRUE),
    avg_no2 = mean(avg_no2, na.rm = TRUE),
    avg_o3 = mean(avg_o3, na.rm = TRUE),

    avg_temperature =
      mean(avg_temperature, na.rm = TRUE),

    avg_humidity =
      mean(avg_humidity, na.rm = TRUE),

    yearly_total_hospital =
      sum(total_hospital, na.rm = TRUE),

    avg_hospital_capacity =
      mean(avg_hospital_capacity, na.rm = TRUE),

    .groups = "drop"
  )
```
```{r top-3-polluted-cities}
# step 3
top3_polluted_cities <- yearly_summary |>

  filter(
    year %in%
      c("2020", "2021", "2022",
        "2023", "2024", "2025")
  ) |>

  group_by(city) |>

  summarise(
    overall_avg_aqi =
      mean(avg_aqi, na.rm = TRUE),

    overall_total_hospital =
      sum(yearly_total_hospital,
          na.rm = TRUE),

    overall_avg_capacity =
      mean(avg_hospital_capacity,
           na.rm = TRUE),

    .groups = "drop"
  ) |>

  arrange(
    desc(overall_avg_aqi)
  ) |>

  slice_head(n = 3)

```

```{r 3-least-polluted-cities}
least3_polluted_cities <- yearly_summary |>

  filter(
    year %in%
      c("2020", "2021", "2022",
        "2023", "2024", "2025")
  ) |>

  group_by(city) |>

  summarise(

    overall_avg_aqi =
      mean(avg_aqi, na.rm = TRUE),

    overall_total_hospital =
      sum(yearly_total_hospital,
          na.rm = TRUE),

    overall_avg_capacity =
      mean(avg_hospital_capacity,
           na.rm = TRUE),

    .groups = "drop"
  ) |>

  arrange(
    overall_avg_aqi
  ) |>

  slice_head(n = 3)
```


```{r Table}
# Combine top 3 and least 3 polluted cities
comparison_cities_names <- c(
  top3_polluted_cities$city,
  least3_polluted_cities$city
)

# Create AQI comparison table for 2020–2025 only

kable(

  yearly_summary |>

    filter(
      city %in% comparison_cities_names,

      year %in%
        c("2020", "2021", "2022",
          "2023", "2024", "2025")
    ) |>

    select(city, year, avg_aqi) |>

    pivot_wider(
      names_from = year,
      values_from = avg_aqi
    ) |>

    rename(City = city),

  caption =
    "Average AQI by Year for the 3 Most and 3 Least Polluted Cities (2020–2025)",

  digits = 2

) |>

  kable_styling(
    bootstrap_options =
      c("striped", "hover", "bordered")
  ) |>

  column_spec(1, bold = TRUE) |>

  row_spec(
    0,
    bold = TRUE,
    background = "#4682B4",
    color = "white"
  )
```

```{r}
  
```

A multiple linear regression model was used to examine the major factor of high AQI, relationship between air pollution and hospital admissions. `Hospital admissions` were treated as the dependent variable, while AQI, PM2.5, PM10, NO2, and O3 were included as predictor variables.

```{r Model-PM25-HospitalAdmission-Correlation}
correlation_model <-lm(hospital_admissions ~ aqi + pm2_5 + pm10 + no2 + o3,
data = airquality_data)

summary(correlation_model)

#Plotting Linear Regg Table

tidy(correlation_model) |>

  rename(
    Variable = term,
    Coefficient = estimate,
    `Standard Error` = std.error,
    `t value` = statistic,
    `p value` = p.value
  ) |>

  kable(
    caption =
      "Linear Regression Results for Hospital Admissions",

    digits = 4
  ) |>

  kable_styling(
    bootstrap_options =
      c("striped", "hover", "bordered")
  ) |>

  row_spec(
    0,
    bold = TRUE,
    background = "#4682B4",
    color = "white"
  )

```

Scatterplots and regression lines were used to visualise the relationship between PM2.5 concentration and hospital admissions across the selected cities.
```{r Plotting the model}

ggplot(airquality_data, aes(x = pm2_5, y = hospital_admissions)) +
  geom_point(alpha = 0.2, color = "#4682B4") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Correlation between Pollutants and Hospital Admissions",
    subtitle = "PM2.5 is the only statistically significant predictor (p < 2e-16)",
    x = "PM2.5",
    y = "Hospital Admissions"
  ) +
  theme_minimal()



```



# Results

# Discussion

# Conclusion

# Recommendations

\newpage

# Potential Results
**Should be used in results**
```{r}
#Using correlation between PM2.5 particles and Hospital admission to provide evidence of relativeness between high pollution and hospital burden

# Filter only the top 3 most polluted cities

# Create pollution group labels

comparison_yearly_data <- comparison_yearly_data |>

  mutate(
    pollution_group = case_when(

      city %in% top3_polluted_cities$city
        ~ "Top 3 Most Polluted",

      city %in% least3_polluted_cities$city
        ~ "3 Least Polluted"
    )
  )

# PM2.5 trend plot

plot1 <- ggplot(
  comparison_yearly_data,

  aes(
    x = year,
    y = avg_pm25,
    color = city,
    linetype = pollution_group,
    group = city
  )
) +

  geom_line(linewidth = 1) +

  geom_point(size = 2) +

  labs(
    title = "PM2.5 Trends by City",
    x = "Year",
    y = "Average PM2.5",
    color = "City",
    linetype = "Pollution Group"
  ) +

  theme_minimal()

# Hospital admissions trend plot

plot2 <- ggplot(
  comparison_yearly_data,

  aes(
    x = year,
    y = yearly_total_hospital,
    color = city,
    linetype = pollution_group,
    group = city
  )
) +

  geom_line(linewidth = 1) +

  geom_point(size = 2) +

  labs(
    title = "Hospital Admissions by City",
    x = "Year",
    y = "Yearly Total Hospital Admissions",
    color = "City",
    linetype = "Pollution Group"
  ) +

  theme_minimal()

# Combine plots

plot1 / plot2 +

  plot_annotation(
    title =
      "Comparison of Pollution and Hospital Admissions",

    subtitle =
      "Top 3 Most vs 3 Least Polluted Cities (2020–2025)"
  )

```


