---
title: "Air Quality and Respiratory Health: An Analysis on Healthcare System Loading"
author:
- name: Somya Verma
  degrees: BAEco(Hons), MBAt
  email: sver0044@student.monash.edu
- name: Daniel Loza
  degrees: AMusA, BBus, BCom(Hons)
- name: Yunhui Fan
  degrees: BSc(Hons), MBAt
- name: Yixiang Xu
  degrees: BManSc, MBAt
phone: (03) 9905 2478
email: BusEco-Econometrics@monash.edu
organization: 6it Consultants
bibliography: references.bib
toc: true
format: report-pdf
---

\newpage

# Executive summary

# Introduction

# Methodology

# Results

# Discussion

# Conclusion

# Recommendations

\newpage

# Introduction 
Question to investigate: **Impact on Hospital Intakes in top 5 most air-polluted cities of world from 2020-2025**

\newpage

# Methodology 
The analysis was conducted using R and several libraries from the tidyverse ecosystem for data cleaning, transformation, and visualisation. Additional packages including lubridate, ggplot2, kableExtra, and patchwork were used for date handling, plotting, and professional table formatting.
```{r loading-library}
library(tidyverse)
library(lubridate)
library(here)
library(ggplot2)
library(knitr)
library(patchwork)
library(kableExtra)

```

The dataset contains daily air quality and healthcare observations from multiple cities between 2020 and 2025. Key variables include Air Quality Index (AQI), PM2.5, PM10, NO2, O3, temperature, humidity, and hospital admissions. The dataset was used to examine whether increasing pollution levels were associated with higher hospital intake rates in heavily polluted cities.
```{r reading-data}
list.files("data")
airquality_data <- read_csv("data/air_quality.csv")
```

To identify the most polluted cities, the average AQI for each city was calculated using grouped summary statistics. Cities were then ranked according to their average AQI, and the top five most polluted cities were selected for further analysis.

Daily observations were aggregated into monthly and yearly averages to reduce short-term fluctuations and improve trend interpretation. Average AQI, PM2.5 concentration, and hospital admissions were calculated for each city and year. These summaries were presented using formatted tables and visualisations.

```{r tidying-data}
airquality_data<- airquality_data |> 
  mutate(
    month = month(date, label = TRUE, abbr= TRUE),
    year = as.factor(year(date))
  )
# Step 1 - identify top 5 most polluted cities
top5_cities <- airquality_data %>%
  group_by(city) %>%
  summarise(avg_aqi = mean(aqi, na.rm = TRUE)) %>%
  arrange(desc(avg_aqi)) %>%
  slice_head(n = 5) %>%
  pull(city)

# Step 2 - yearly summary for top 5 cities

#(With Months)
yearly_data <- airquality_data %>%
  filter(city %in% top5_cities,
         year %in% as.factor(2020:2025)) %>%
  group_by(city, year, month) %>%
  summarise(
    avg_aqi = mean(aqi, na.rm = TRUE),
    avg_pm25 = mean(pm2_5, na.rm = TRUE),
    avg_hospital = mean(hospital_admissions, na.rm = TRUE),
    .groups = "drop"
  )

#(Without month)
yearly_summary <- yearly_data %>%
  group_by(city, year) %>%
  summarise(
    avg_aqi = mean(avg_aqi, na.rm = TRUE),
    avg_pm25 = mean(avg_pm25, na.rm = TRUE),
    avg_hospital = mean(avg_hospital, na.rm = TRUE),
    .groups = "drop"
  )

kable(yearly_summary %>%
        select(city, year, avg_aqi) %>%
        pivot_wider(names_from = year, values_from = avg_aqi) %>%
        rename(City = city),
      caption = "Average AQI by City and Year (2020-2025)",
      digits = 2) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(0, bold = TRUE, background = "SteelBlue", color = "white")
```

A multiple linear regression model was used to examine the relationship between air pollution and hospital admissions. `Hospital admissions` were treated as the dependent variable, while AQI, PM2.5, PM10, NO2, and O3 were included as predictor variables.

```{r Model-PM25-HospitalAdmission-Correlation}
correlation_model <-lm(hospital_admissions ~ aqi + pm2_5 + pm10 + no2 + o3,
data = airquality_data)

summary(correlation_model)
```

Scatterplots and regression lines were used to visualise the relationship between PM2.5 concentration and hospital admissions across the selected cities.
```{r Plotting the model}

ggplot(airquality_data, aes(x = pm2_5, y = hospital_admissions)) +
  geom_point(alpha = 0.2, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Correlation between Pollutants and Hospital Admissions",
    subtitle = "PM2.5 is the only statistically significant predictor (p < 2e-16)",
    x = "PM2.5",
    y = "Hospital Admissions"
  ) +
  theme_minimal()
```


# Result 
**Should be used in results**
```{r}
#Using correlation between PM2.5 particles and Hospital admission to provide evidence of relativeness between high pollution and hospital burden

plot1 <- ggplot(yearly_summary, aes(x = year, y = avg_pm25, color = city, group = city)) +
  geom_line() +
  geom_point() +
  labs(
    title = "PM2.5 Trends by City",
    x = "Year",
    y = "Average PM2.5",
    color = "City"
  ) +
  theme_minimal()

plot2 <- ggplot(yearly_summary, aes(x = year, y = avg_hospital, color = city, group = city)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Hospital Admissions by City",
    x = "Year",
    y = "Average Hospital Admissions",
    color = "City"
  ) +
  theme_minimal()

plot1 / plot2 +
  plot_annotation(
    title = "PM2.5 and Hospital Admissions Trends",
    subtitle = "Top 5 Most Polluted Cities (2020-2025)"
  )
```

```{r}

aqi_long <- yearly_summary %>%
  select(city,year,avg_aqi)

bar_aqi <- ggplot(aqi_long,aes(x=year,y=avg_aqi,fill=city))+
  geom_col(position = position_dodge(width=0.9),width=0.8)+
  labs(title="Annual Average AQI across five cities (2020-2025)",x="Year",y="Average AQI",fill="City")+
  geom_vline(xintercept = "2020", linetype="dashed",color="black")+
  theme_minimal()

bar_aqi
```

**this is an extra filteration we can use in result**

```{r extra-info}
# Identify top 5 most polluted months per city per year
polluted_months <- yearly_data %>%
  group_by(city, year) %>%
  slice_max(avg_aqi, n = 5) %>%
  ungroup()

# Most polluted year and month per city
city_pollution_summary <- yearly_data %>%
  group_by(city) %>%
  summarise(
    avg_aqi = mean(avg_aqi, na.rm = TRUE),
    most_polluted_year = year[which.max(avg_aqi)],
    most_polluted_month = month[which.max(avg_aqi)]
  )
```
```{r}
month_pollute_plot <- ggplot(polluted_months,aes(x=month,y=avg_aqi,fill=city))+
  geom_col(position="dodge",alpha=0.7)+
  facet_wrap(~city,scales="free_y",ncol=2)+
  labs(title="Top 5 highest-AQI months per city each year",x="Month",y="Average AQI")+
  theme_minimal()

month_pollute_plot
```

This section presents descriptive pollution trends across five selected high-pollution cities, multivariate regression outcomes quantifying pollutant-hospitalisation links, and supplementary monthly extreme-pollution analysis, supported by tabular summaries and graphical visualisations. All analysis covers 2020 to 2025 daily observational data aggregated into monthly and annual averages to mitigate short-run random fluctuations.

## Annual Air Quality Descriptive Findings across Top 5 Cities

Table 1 summarises annual average AQI values for Beijing, Cairo, Delhi, Los Angeles and São Paulo, the five cities filtered by long-run average AQI as the most polluted globally, while Figure 1 displays yearly AQI distribution via grouped bar charts.
A distinct abnormal pollution trough was observed universally in 2020 across all five cities, consistent with COVID-19 pandemic restrictions: nationwide factory suspension, travel curbs and private vehicle reduction sharply cut industrial and traffic pollutant emissions. Most prominently, São Paulo recorded the highest baseline annual AQI of 392.00 in 2020 before a dramatic sustained decline to 190.08 by 2025, representing a near 52% drop in average pollution over six years. By contrast, Beijing maintained stable high AQI levels throughout the sample period, rising gradually from 272.85 (2020) to 289.64 (2025), with limited pollution improvement. Cairo exhibited the largest year-on-year volatility, peaking at 417.60 AQI in 2024 followed by a steep fall to 193.14 in 2025. Delhi and Los Angeles saw mild, modest fluctuations within a narrow AQI band between roughly 215 and 265 across all years with no clear long-term rising or falling trend.
Figure 2 presents paired time-series line graphs tracking annual average PM2.5 concentrations and corresponding annual hospital admission volumes separately for each city from 2020–2025. Visually synchronous co-movement is evident: years with falling PM2.5 concentrations consistently correspond to lower average hospital admissions, most notably for São Paulo, whose sharp post-2020 PM2.5 reduction coincides with a persistent downtrend in hospital intake counts. The vertical dashed marker for 2020 further confirms the pandemic-driven synchronous drop of both pollution and hospital admissions across all sample cities.
